---
title: "EDI Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDI Workflows}
  %\VignetteEngine{knitr::rmarkdown}
---

The following demonstrates how to grab files from the EDI Repository. These functions offer similar services to those from the 'EDIUtils' package but are opinionated to work similarly to other workflows in 'deltadata'.

```{r setup}
library(deltadata)
```

## Reading in entities from EDI

The `getEDI()` function is a one-stop shop function that allows us to download entities from an EDI repository. To get started, provide the function with a URL to the data package of interest.

```{r}
# Data package of the CDFW IEP SLS Survey
possibleEntities <- getEDI("https://portal.edirepository.org/nis/mapbrowse?packageid=edi.534.9")
```

When only a URL is provided, the function will return a list of entities available to download. The `name` is the file name of the entity, `extension` is the file extension, `size` is the file size, and `description` is a short description of the entity provided by the creators. This printed table contains only a subset of the entire data.frame, which can be accessible if you assign the function to a variable.

```{r}
# Full data.frame has more information than what is printed
head(possibleEntities)
```

To download file(s), we can provide the values of interest in the `name` column to the function.

```{r}
entities <- getEDI("https://portal.edirepository.org/nis/mapbrowse?packageid=edi.534.9", 
                   files = c("Catch.csv", "SLSTables.rds", "SLS_Metadata.pdf"))
```

## Files are stored differently

The function download and stores the entities based on what the file extension. Specifically, '.csv' and '.rds' files are read directly into R, while all other file types are downloaded to the temporary folder and a link to the file is provided. 

```{r}
# csv files are read in directly
head(entities$Catch.csv)
```

```{r}
# rds files are read in directly, outputted as a list
lapply(entities$SLSTables, head)
```

```{r}
# All other file types are temporarily downloaded and a file path provided
entities$SLS_Metadata
```

You can use `shell.exec()` to open any downloaded files.

```{r, eval=FALSE}
shell.exec(entities$SLS_Metadata)
```

